---
title: "Exploratory Data Analysis and Cleaning for RITA Flight Data"
author: "Gergana Ilieva"
date: "June 23, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##About

This dataset is comprised by the US airlines flight delays & performance. It 
includes all flights from all carriers from January 2005 to January 2016. 
 
[id]: http://www.transtats.bts.gov/OT_Delay/ot_delaycause1.asp?display=download&pn=0&month=11&year=2014
[The data can be downloaded from the RITA site.][id]

##Quick inspection of the data

Let's start by quickly reviewing the structure and summary of our dataset.

```{r echo=FALSE, Quick_Summary}
flights <- read.csv('data/413498804_42016_3325_airline_delay_causes.csv')
str(flights)
summary(flights)
head(flights)
```

##Data cleaning

```{r echo=FALSE}
library(dplyr)
flights$date <- as.Date(paste(flights$year, flights$X.month, 1, sep = '-'),
                        format = "%Y-%m-%d")
summary(flights$date)

```

If we want to quickly check how many distinct carriers do we have in our data set:
```{r echo = FALSE}
nrow(table(flights$carrier))
```

In order to make a new summary table & get rid of the NA values (in case of any)
we apply the following code

```{r echo = FALSE}
flights_new = flights %>%
  group_by(date, year, carrier_name) %>%
  summarize (arrivals = sum(arr_flights),
             delayed = sum(arr_del15),
             cancelled = sum(arr_cancelled),
             diverted = sum(arr_diverted)) %>%
  transform(ontime = 1 - delayed/arrivals)

flights_new <- flights_new[complete.cases(flights_new),]
str(flights_new)
```

##Exploratory plot
```{r echo = FALSE}
library(ggplot2)
ggplot(data = flights_new, aes(x = date, y = ontime)) + 
  geom_line(aes(color = carrier_name))
```

##Find airlines to subset data (aggregation by carrier)
```{r echo = FALSE}
aggrr <- flights_new %>%
  group_by(carrier_name) %>%
  summarize(mnt_avg = mean(arrivals),
            arrivals = sum(arrivals))

sel_carriers <- subset(aggrr, mnt_avg >= quantile(mnt_avg, 0.81))$carrier_name
sel_carriers
```

##Reshape data
```{r echo = FALSE}
df <- subset(flights_new, is.element(carrier_name, sel_carriers)) %>%
  group_by(year, carrier_name) %>%
  summarize(arrivals = sum(arrivals),
            delayed = sum(delayed),
            cancelled = sum(cancelled), 
            diverted = sum(diverted)) %>%
  transform(ontime = 1 - delayed/arrivals)
df <- df[complete.cases(df),]
head(df)
```

##Plot again
```{r echo = FALSE}
library(ggplot2)
summary(flights$year)

p1 <- ggplot(data = df,
       aes(x = year, y = ontime)) +
  geom_line(aes(color = carrier_name)) +
  scale_x_continuous(limits=c(2005, 2015), breaks=c(2005:2015))

p1

p2 <- ggplot(data = df,
       aes(x = year, y = arrivals)) +
  geom_line(aes(color = carrier_name)) +
  scale_x_continuous(limits=c(2005, 2015), breaks=c(2005:2015))

p2
```

##Export new data file
```{r echo = FALSE}
write.csv(df, file="data.csv", row.names=FALSE)
```

